---
title: "Matrix Algebra"
author: "Gaston Sanchez"
date: "November 7, 2016"
output: pdf_document
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(knitr)
read_chunk('matrix-practice-chunks.R')
```


# Basic Vector and Matrix manipulations in R

If you have no or minimal experience in R working with matrices, please go over
this section. If you are already familiar with this material, skip to the next
section.

Consider the following vector `x`:
```{r vectorx}
```

- Use the vector `x` as input of the function `matrix()` to create the
following matrix:
```{r matrix1, echo = FALSE, comment=''}
```

- Using `x` and `matrix()`, how would you generate a matrix like this:
```{r matrix2, echo = FALSE, comment=''}
```

- Use `diag()` to create the following identity matrix $\mathbf{I}_n$ of 
dimensions $n \times p$ = (5, 5):
```{r identity5, echo = FALSE, comment=''}
```

- Consider the following vectors `a1, a2, a3`:
```{r vectors-a}
```

- Column-bind the vectors `a1, a2, a3` to form this matrix:
```{r matrix3, echo = FALSE, comment=''}
```

- Now row-bind the vectors `a1, a2, a3` to form this matrix:
```{r matrix4, echo = FALSE, comment=''}
```

- Using matrix functions like transpose `t()` and product `%*%`, write a function 
`vnorm()` that computes the Euclidean norm (i.e. length) of a vector: 
$\| \mathbf{x} \| = \sqrt{\mathbf{x^\mathsf{T} x}}$.

```{r eval = FALSE}
# test vnorm() with 1:5
v <- 1:5
vnorm(v)
```

- Given the vector `v <- 1:5`, use `vnorm()` to create a vector `u` such that
`u` is of unit norm: $\| \mathbf{u} \| = 1$.

- Write a function `is_square()` to check whether the provided matrix is a 
square matrix.

- Write a function `mtrace()` to compute the trace of a square matrix. Use 
your `is_square()` function to make sure that the input is a square matrix.
The trace is defined as the sum of the elements on the diagonal:
$tr(\mathbf{A}) = \sum_{i=1}^{n} a_{ii}$

- Given two square matrices $\mathbf{A}$ and $\mathbf{B}$, verify that your 
function `tr()` is a linear mapping:
    + $tr(\mathbf{A+B}) = tr(\mathbf{A}) + tr(\mathbf{A})$
    + $tr(c \mathbf{A}) = c \times tr(\mathbf{A})$, where $c$ is a scalar
    
- Trace of products: Given two matrices $\mathbf{X}$ and $\mathbf{Y}$ of 
adequate size, verify that:
$$
tr(\mathbf{X^\mathsf{T} Y}) = tr(\mathbf{X Y^\mathsf{T}}) = tr(\mathbf{Y^\mathsf{T} X}) = tr(\mathbf{Y X^\mathsf{T}})
$$

- Cross-products: Use `proc.time()` or `system.time()` to compare the 
`t(X) %*% X` against `crossprod(X)`

- Cross-products: Use `proc.time()` or `system.time()` to compare the 
`X %*% t(X)` against `tcrossprod(X)`


# Transformation and Scaling Operations

In this section you will be using the built in data.frame `mtcars` to practice
transforming and scaling operations:

```{r}
head(mtcars)
```

- Create a matrix `M` with variables `mpg`, `disp`, `hp`, `drat`, and `wt`.
- Use `apply()` to compute the vector containing the means of the columns in `M`
- Compute a matrix `Mc` of mean-centered data applying the function `scale()`
on `M` (do NOT `scale = TRUE`).
- Confirm that variables in `Mc` are mean-centered by calculating the vector
of column-means
- Use the function `sweep()` to mean-center `M` by _sweeping out_ the vector of
column means. Compare this result with `Mc` (you should get the same values).
- Compute a vector of column maxima.
- Use `sweep()` to scale the columns by dividing by the column maxima.
- Compute a matrix in which all columns are scaled to have minimum = 0 and
maximum = 1

- Write a function `dummify()` that takes a factor or a character vector, and 
which returns a matrix with dummy indicators. Assuming that the factor has $k$ categories (i.e. $k$ `levels`), include an argument `all` that lets you specify 
whether to return $k$ binary indicators, or $k-1$ indicators. You should
be able to call `dummify()` like this:
```{r eval = FALSE}
cyl <- factor(mtcars$cyl)
# all categories
CYL1 <- dummify(cyl, all = TRUE)
# minus one category
CYL2 <- dummify(cyl, all = FALSE)
```

- Write a function `crosstable()` that takes two factors, and which returns 
a cross-table between those factors. To create this function you should use 
your function `dummify()` to compute two dummy matrices, and then multiple them.
```{r eval = FALSE}
cyl <- factor(mtcars$cyl)
gear <- factor(mtcars$gear)
xtb <- crosstable(cyl, gear)
```

You should get a table like this:
```
   3 4 5
4  1 8 2
6  2 4 1
8 12 0 2
```


\newpage

# Solutions

```{r}
<<matrix1>>
<<matrix2>>
<<identity5>>
<<matrix3>>
<<matrix4>>
```

```{r}
<<vnorm>>
<<unitnorm>>
<<is_square>>
<<mtrace>>
<<lineartrace>>
<<prodtrace>>
<<crossprod>>
<<tcrossprod>>
```

```{r}
<<transformations>>
```

```{r}
<<dummify>>
```

```{r}
<<crosstable>>
```

