---
runtime: shiny
output: html_document
---

```{r echo=FALSE}
library(babynames)
library(ggplot2)
library(stringr)
```

## Shiny Document

Please select a name to explore.

```{r echo=FALSE}
textInput(inputId = "name", label = "Name:", value = "Gaston")
names <- reactive(subset(babynames, name == input$name))

# another option
#names <- reactive(subset(babynames, name == str_to_title(input$name)))
```


Here is a graph of the number of people named `r renderText(input$name)`.

```{r echo=FALSE, fig.height=3, fig.width=5}
renderPlot(
  ggplot(data = names(), aes(x = year, y = n)) +
    geom_line(aes(color = sex), size = 1.5) + 
    theme_bw()
)
```

`r renderText(input$name)` was more popular than ever in 
`r renderText(names()$year[which.max(names()$n)])`. 
That year there were `r renderText(names()$n[which.max(names()$n)])` babies named `r renderText(input$name)`.
So far, `r renderText(sum(names()$n))` people have been named `r renderText(input$name)`. 

`r hr()`

Here's an example of `renderPrint()`
`r renderPrint(paste('Selected name is:', input$name))`

`r hr()`

Here's an example of `renderTable()` showing the last rows in the subset dataframe `babynames`:
`r renderTable(tail(names()))`

