---
title: "Stat 243: Lab, Monday Oct-17"
header-includes: \usepackage{float}
output: pdf_document
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(xtable)
source("die-roll-s3class.R")
```


# OOP in R: Rolling a Die

The purpose of this lab is to practice Object-Oriented Programming (OOP) using
either S3 or S4 classes in R.

The goal is to program two classes of objects: a regular `"die"` with 
six sides, and an object `"roll"` (i.e. the rolls 
of a `"die"` multiples times).

- Use an `.R` script file to write code for the class objects and methods
- If you are already familiar with S3 classes, try using S4 classes
- If you are new to both S3 and S4 classes, choose one of them

-----

# Object `"die"`

The object `"die"` should have two attributes: 

- `sides`: vector with numbers 1, 2, 3, 4, 5, 6.
- `prob`: vector of probabilities for each side

For the `"die"` object write:

- a constructor function `die()` that creates a _fair_ die by default
    + make sure that the argument `prob` has correct probability values
- a `"print"` method that displays the sides and the associated probabilities

You should be able to use `die()` like this:
```{r eval = TRUE}
# create a fair die by default
fair_die <- die()
fair_die
```


```{r}
# create a loaded die
loaded_die <- die(prob = c(0.075, 0.1, 0.125, 0.15, 0.20, 0.35))
loaded_die
```

-----

# Object `"roll"`

The object `"roll"` should have these attributes: 

- `rolls`: vector with outputs of the rolls
- `sides`: vector with numbers 1, 2, 3, 4, 5, 6.
- `prob`: vector of probabilities for each side

For the `"roll"` object write:

- a constructor function `roll()` that _rolls_ a `"die"` a given number of `times`
    + give `times` a default value of 1
- a `print()` method that displays just the rolls
- a `summary()` method that displays a table of frequencies: both the counts 
and proportions of each side of the rolled `die` 
- a `plot()` method using a barchart of frequencies (relative frequencies of
`1`'s, `2`'s, `3`'s, `4`'s, `5`'s, and `6`'s) 
- an extraction method `"["` to extract the value of a given roll
- a replacement method `"[<-"` to replace the value of a given roll
- an addition `"+"` method to add more rolls


You should be able to use `roll()` and its methods like this:

```{r eval = TRUE}
# roll fair die
fair500 <- roll(fair_die, times = 500)

# summary method
summary(fair500)

# retrieving roll in position 100
fair500[100]

# replacing last roll
fair500[500] <- 1

# adding 100 rolls
fair600 <- fair500 + 100

# plot method
plot(fair500, 500)
```


## De Mere's problem

Use your objects `"die"` and `"roll()"` to simulate a series of
1000 games for the famous _Chevalier De Mere's_ dice problems:

- [One gambling problem that launched modern probability theory](https://introductorystats.wordpress.com/2010/11/12/one-gambling-problem-that-launched-modern-probability-theory/) (by Dan Ma).

The first problem involves computing the probability of getting at least one 
"6" in four rolls of a die. This probability can be computed analytically as:

$$
1 - (5/6)^4
$$

The other problem involves computing the probability of of getting at least
two "6" in 24 rolls of a pair of dice. This probability can be computed 
analytically as:

$$
1 - (35/36)^{24}
$$

The goal is to use your `roll()` function to simulate a series of
1000 games for both of De Mere's problems:

- one series of 1000 games should involve rolling a die four times, and then
count the number of games in which there is at least one 6.
- the other series of 1000 games should involve rolling a pair of dice 24 times, 
and then count the number of games in which there is at least one double 6.


## Testing whether a die is fair

Here's one more challenge. The idea is to write R code, preferably a function,
such that given a vector of outputs from rolling a die, it can be determined
whether the die is fair.

One way to approach this problem is by using Pearson's Chi-square statistic:

$$
\chi^2 = \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i} = N \sum_{i=1}^{n} \frac{(O_i/ N - p_i)^2}{p_i}
$$

where:

- $n$ is the number of sides (i.e. 6)
- $O_i$ is the observed frequency of side $i$, $i = 1, 2, 3, 4, 5, 6$
- $E_i = N p_i$ is the expected frequency of side $i$
- $p_i$ is the theoretical relative frequency ($1/6$ assuming a fair roll)
- $N$ is the total number of rolls

When testing for a fair six-sided die, the statistic $\chi^2$ asymptotically 
approaches a chi-square distirbution with degrees of freedom $n - 1 = 5$.

You can generate a loaded die, and roll it 1000 times:
```{r}
set.seed(3625)
loaded_die <- die(prob = c(0.075, 0.1, 0.125, 0.15, 0.20, 0.35))
load_rolls <- roll(loaded_die, 1000)
```

Try implementing the chi-square statistic and compute a hypothesis test
to see whether the rolls in `load_rolls` support the null hypothesis of a fair
die.


```{r eval = FALSE, echo = FALSE}
chi_square(load_rolls)

pchisq(330.728, df = 5, lower.tail = FALSE)
pchisq(7.192, df = 5, lower.tail = FALSE)
```
