---
title: "Stat 243"
subtitle: "Base Graphics (part 1)"
author: "Gaston Sanchez"
date: "Creative Commons Attribution 4.0 International License"
output: 
  beamer_presentation:
    includes:
      in_header: header.tex
---

```{r, setup, include=FALSE}
library(knitr)
# smaller font size for chunks
knitr::opts_chunk$set(size = 'footnotesize')
options(width=78)
```

# 

\begin{center}
\Huge{\hilit{Graphics}}
\end{center}


# References

Some Resources

- __R Graphics__ by Paul Murrell
- __R Graphics Cookbook__ by Winston Chang
- __ggplot2: Elegant Graphics for Data Analysis__ by Hadley Wickham
- __R Graphs Cookbook__ by Hrishi Mittal
- __Graphics for Statistics and Data Analysis with R__ by Kevin Keen



# Two main graphic systems

\centerline{\mdlit \Large 2 main graphics systems}

\bigskip
\centerline{\Large "graphics" \quad \& \quad "grid"}


# Basics of Graphics in R

## Graphics Systems

- `"graphics"` and `"grid"` are the two main graphics systems in R
- `"graphics"` is the _traditional_ system, also referred to as _base graphics_
- `"grid"` prodives low-level functions for programming plotting functions


# Basics of Graphics in R

## Graphics Engine

- Underneath `"graphics"` and `"grid"` there is the package `"grDevices"`
- `"grDevices"` is the graphics __engine__ in R
- It provides the graphics devices and support for colors and fonts


# 

\begin{center}
\includegraphics[width=8cm]{graphic_systems.pdf}
\end{center}


# Basics of Graphics in R

## Package `"graphics"`

The package `"graphics"` is the traditional system; it provides functions for 
complete plots, as well as low-level facilities. 

\bigskip

Many other graphics packages are built on top of graphics like `"maps"`, `"diagram"`, `"pixmap"`, and many more.


# Understanding Graphics in R

## Package `"grid"`

The `"grid"` package does not provide functions for drawing complete plots.

\bigskip

`"grid"` is not used directly to produce statistical plots. Instead, it is 
used to build other graphics packages like `"lattice"` or `"ggplot2"`.


# 

\begin{center}
\Huge{\hilit{Traditional (Base) Graphics}}
\end{center}


# Base Graphics in R

## Types of graphics functions

Graphics functions can be divided into two main types:

- __high-level__ functions produce complete plots, e.g. `barplot()`, 
`boxplot()`, `dotchart()`
- __low-level__ functions add further output to an existing plot, e.g. 
`text()`, `points()`, `legend()`


# The `plot()` function

- `plot()` is the most important high-level function in traditional graphics
- The first argument to `plot()` provides the data to plot
- The provided data can take different forms: e.g. vectors, factors, matrices, 
data frames.
- To be more precise, `plot()` is a generic function
- You can create your own `plot()` method function


# Basic Plots with `plot()`

In its basic form, we can use `plot()` to make graphics of:

- one single variable
- two variables
- multiple variables


#

\begin{center}
\Huge{\hilit{Plots of One Variable}}
\end{center}


# High-level graphics of a single variable

| Function | Data      | Description |
|----------| ----------|-------------|
| `plot()` | numeric   | scatterplot |  
| `plot()` | factor    | barplot     |
| `plot()` | 1-D table | barplot     |

\bigskip

`numeric` can be either a vector or a 1-D array 
(e.g. row or column from a matrix) 


# Examples

```{r plot_single1, eval=FALSE, echo=-c(1,14)}
op <- par(mfrow = c(1, 3))
# plot numeric vector
num_vec <- (c(1:10))^2
plot(num_vec)

# plot factor
set.seed(4)
abc <- factor(sample(c('A', 'B', 'C'), 20, 
                     replace = TRUE))
plot(abc)

# plot 1D-table
abc_table <- table(abc)
plot(abc_table)
par(op)
```


#

```{r plot_single1, fig.width=7, fig.height=4, out.width='.9\\linewidth', out.height='.4\\linewidth', fig.align='center', echo=FALSE}
```


# More high-level graphics of a single variable

\begin{center}
 \begin{tabular}{l l l}
  \hline
   Function & Data & Description \\
  \hline
  \code{barplot()} & numeric & barplot \\
  \code{pie()} & numeric & pie chart  \\
  \code{dotchart()} & numeric & dotplot \\
   & & \\
  \code{boxplot()} & numeric & boxplot \\
  \code{hist()} & numeric & histogram  \\
  \code{stripchart()} & numeric & 1-D scatterplot \\
  \code{stem()} & numeric & stem-and-leaf plot \\
  \hline
 \end{tabular}
\end{center}


# Kernel Density Curve

- Surprisingly, R does not have a specific function to plot density curves
- R does have the `density()` function which computes a kernel density estimate
- We can pass a `"density"` object to `plot()` in order to get a density curve.



# Kernel Density Curve

```{r fig.width=4, fig.height=4, out.width='.5\\linewidth', out.height='.5\\linewidth', fig.align='center'}
# kernel density curve
dens <- density(num_vec)
plot(dens)
```


#

\begin{center}
\Huge{\hilit{Plots of Two Variables}}
\end{center}


# High-level graphics of two variables

\begin{center}
 \begin{tabular}{l l l}
  \hline
   Function & Data & Description \\
  \hline
  \code{plot()} & numeric, numeric & scatterplot \\  
  \code{plot()} & numeric, factor & stripcharts \\
  \code{plot()} & factor, numeric & boxplots \\
  \code{plot()} & factor, factor & spineplot \\
  \code{plot()} & 2-column numeric matrix & scatterplot \\  
  \code{plot()} & 2-column numeric data.frame & scatterplot \\  
  \code{plot()} & 2-D table & mosaicplot \\
  \hline
 \end{tabular}
\end{center}


# Plots of two variables

```{r plot_twovars, eval=FALSE, echo=-c(1,13)}
op <- par(mfrow = c(1, 3))
# plot numeric, numeric
plot(iris$Petal.Length, iris$Sepal.Length)

# plot numeric, factor
plot(iris$Petal.Length, iris$Species)

# plot factor, numeric
plot(iris$Species, iris$Petal.Length)

# plot factor, factor
plot(iris$Species, iris$Species)
par(op)
```


# Plots of two variables

```{r fig.width=5, fig.height=4, out.width='.7\\linewidth', out.height='.6\\linewidth', fig.align='center'}
# plot numeric, numeric
plot(iris$Petal.Length, iris$Sepal.Length)
```


# Plots of two variables

```{r fig.width=5, fig.height=4, out.width='.7\\linewidth', out.height='.6\\linewidth', fig.align='center'}
# plot numeric, factor
plot(iris$Petal.Length, iris$Species)
```


# Plots of two variables

```{r fig.width=5, fig.height=4, out.width='.7\\linewidth', out.height='.6\\linewidth', fig.align='center'}
# plot factor, numeric
plot(iris$Species, iris$Petal.Length)
```


# Plots of two variables

```{r fig.width=6, fig.height=6, out.width='.7\\linewidth', out.height='.6\\linewidth', fig.align='center'}
# plot factor, factor
plot(iris$Species, iris$Species)
```


# More high-level graphics of two variables

\footnotesize{
\begin{center}
 \begin{tabular}{l l l}
  \hline
   Function & Data & Description \\
  \hline
  \code{sunflowerplot()} & numeric, numeric & sunflower scatterplot \\
  \code{smoothScatter()} & numeric, numeric & smooth scatterplot \\
   & & \\
  \code{boxplot()} & list of numeric & boxplots \\
  \code{barplot()} & matrix & stacked / side-by-side barplot  \\
  \code{dotchart()} & matrix & dotplot \\
   & & \\
  \code{stripchart()} & list of numeric & stripcharts \\
  \code{spineplot()} & numeric, factor & spinogram  \\
  \code{cdplot()} & numeric, factor & conditional density plot  \\
   & & \\
  \code{fourfoldplot()} & 2x2 table & fourfold display \\
  \code{assocplot()} & 2-D table & association plot \\  
  \code{mosaicplot()} & 2-D table & mosaic plot \\
  \hline
 \end{tabular}
\end{center}
}


#

\begin{center}
\Huge{\hilit{Base Graphics}}
\end{center}


# Graphics in R

## Traditional Graphics

- R `"graphics"` follows a static, "painting on canvas" model.
- Graphics elements are drawn, and remain visible until painted over.
- For dynamic and/or interactive graphics, base R graphics are limited.


# Traditional Graphics}

## Traditional graphics model

In the traditional model, we create a plot by first calling a high-level 
function that creates a complete plot, and then we call low-level functions 
to add more output if necessary


# Dataset `mtcars`

```{r}
head(mtcars)
```


# Scatter plot

```{r mtcars1, fig.width=5, fig.height=4, out.width='.8\\linewidth', out.height='.55\\linewidth', fig.align='center'}
# simple scatter-plot
plot(mtcars$mpg, mtcars$hp)
```


# Axis Labels

```{r mtcars2, eval = FALSE}
# xlab and ylab
plot(mtcars$mpg, mtcars$hp, 
     xlab = "miles per gallon",
     ylab = "horsepower")
```


# Axis Labels

```{r mtcars2, fig.width=5, fig.height=4, out.width='.85\\linewidth', out.height='.6\\linewidth', fig.align='center', echo = FALSE}
```


# Title and Subtitle

```{r mtcars3, eval = FALSE}
# title and subtitle
plot(mtcars$mpg, mtcars$hp, 
     xlab = "miles per gallon",
     ylab = "horsepower", 
     main = "Simple Scatterplot",
     sub = 'data mtcars')
```


# Title and Subtitle

```{r mtcars3, fig.width=5, fig.height=4, out.width='.85\\linewidth', out.height='.7\\linewidth', fig.align='center', echo = FALSE}
```


# x and y coordinate ranges

```{r mtcars4, eval = FALSE}
# 'xlim' and 'ylim'
plot(mtcars$mpg, mtcars$hp,
     xlim = c(10, 35), 
     ylim = c(50, 400),
     xlab = "miles per gallon",
     ylab = "horsepower", 
     main = "Simple Scatterplot",
     sub = 'data mtcars')
```


# x and y coordinate ranges

```{r mtcars4, fig.width=5, fig.height=4, out.width='.85\\linewidth', out.height='.7\\linewidth', fig.align='center', echo = FALSE}
```


# Points

```{r mtcars5, eval = FALSE}
# character expansion 'cex' and 'point character'
plot(mtcars$mpg, mtcars$hp,
     xlim = c(10, 35), 
     ylim = c(50, 400),
     cex = 1.5, 
     pch = 1,
     xlab = "miles per gallon",
     ylab = "horsepower", 
     main = "Simple Scatterplot",
     sub = 'data mtcars')
```


# Points

```{r mtcars5, fig.width=5, fig.height=4, out.width='.85\\linewidth', out.height='.7\\linewidth', fig.align='center', echo = FALSE}
```


# Point symbols `pch` available in R

```{r points_pch, size='scriptsize', fig.width=4, fig.height=4, out.width='.6\\linewidth', out.height='.6\\linewidth', fig.align='center', echo=FALSE}
x <- rep(1:5, times = 5)
y <- rep(5:1, each = 5)
op <- par(mar = c(1,1,1,1))
plot(x, y, pch = 1:25, axes = FALSE, xlab = '', ylab = '',
     ylim = c(0.5, 5), cex = 1.5)
text(x, y, pch = 1:25, labels = 1:25, pos = 1, offset = 0.8)
par(op)
```


# Point Character

```{r fig.width=5, fig.height=4, out.width='.85\\linewidth', out.height='.6\\linewidth', fig.align='center', size='scriptsize'}
# 'pch' can be any character
plot(mtcars$mpg, mtcars$hp, pch = "@")
```


# Point Colors

```{r fig.width=5, fig.height=4, out.width='.85\\linewidth', out.height='.6\\linewidth', fig.align='center', size='scriptsize'}
# color argument 'col'
plot(mtcars$mpg, mtcars$hp, pch=19, col="blue", cex=1.2)
```


# Coloring Point Symbols

- the `col` argument can be used to color symbols
- symbols 21 through 25 can additionally have
their interiors filled by using the `bg` (background) argument


# Coloring Point symbols

```{r points_cols, size='scriptsize', fig.width=4, fig.height=4, out.width='.6\\linewidth', out.height='.6\\linewidth', fig.align='center', echo=FALSE}
x <- rep(1:5, times = 5)
y <- rep(5:1, each = 5)
cols <- c("tomato", "orange", "green3", "blue", "purple")
op <- par(mar = c(1,1,1,1))
plot(x, y, pch = 1:25, axes = FALSE, xlab = '', ylab = '',
     ylim = c(0.5, 5), col = cols, bg = cols, cex = 2)
text(x, y, pch = 1:25, labels = 1:25, pos = 1, offset = 0.8)
par(op)
```


# Drawing Error Bars

Randomly generate 5 groups of 20 observations each.

```{r}
groups <- rep(1:5, each = 20)
means <- c(5, 7, 6, 9, 8)
y <- rnorm(100) + rep(means, each = 20)
```

Display the data is with parallel boxplots.

```{r group-boxplots, eval = FALSE}
boxplot(y ~ groups, main = "Simulated Grouped Data")
```


# Grouped Box-plots

```{r group-boxplots, fig.width=6, fig.height=4, out.width='.9\\linewidth', out.height='.6\\linewidth', fig.align='center', echo=FALSE}
```


# Error bars

Rather than using boxplots, we can create a new plot that displays the data 
values and shows the mean with error bars.

```{r error-bars, eval = FALSE}
ybar <- sapply(split(y, groups), mean)
ysd <- sapply(split(y, groups), sd)
plot(groups, y, xlim = c(1, 5.25), 
     xlab = "Group", ylab = "Values",
     main = "Data with Means and Error Bars")
points(1:5 + .25, ybar, pch=19)
arrows(1:5 + .25, ybar - 2 * ysd/sqrt(20),
       1:5 + .25, ybar + 2 * ysd/sqrt(20),
       code = 3, angle = 90, length = .1)
```


# Error bars

```{r error-bars, fig.width=6, fig.height=4, out.width='.9\\linewidth', out.height='.6\\linewidth', fig.align='center', echo=FALSE}
```


# Adding some jitter

Notice that it is hard to see the individual data points because of overplotting.
This can be fixed (adding a small amount of random variation in the `x` direction).

```{r jitter, eval = FALSE}
plot(groups + runif(100, -.05, .05), y, 
     xlim = c(1, 5.25), 
     xlab = "Group", ylab = "Values",
     main = "Data with Means and Error Bars")
points(1:5 + .25, ybar, pch=19)
arrows(1:5 + .25, ybar - 2 * ysd/sqrt(20),
       1:5 + .25, ybar + 2 * ysd/sqrt(20),
       code = 3, angle = 90, length = .1)
```


# Adding jitter

```{r jitter, fig.width=6, fig.height=4, out.width='.9\\linewidth', out.height='.6\\linewidth', fig.align='center', echo=FALSE}
```

