---
title: "Getting Started with cointoss"
author: "Gaston Sanchez"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cointoss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(cointoss)
```

## Coin Tossing

The package `"cointoss"` is a minimal implementation for simulating tossing
a coin a given number of times, and to visualize the relative frequencies
in such a series of tosses.


## Creating a `coin`

The first step is to create a `"coin"` object with the function `coin()`:

```{r}
mycoin <- coin()
mycoin
```

By default, `coin()` creates a _fair_ coin with standard sides `heads`, and 
`tails`, and with probability values for each side of 0.5.

You can change the defatuls with arguments `sides` and `prob`:

```{r}
loaded <- coin(c('HEADS', 'tails'), c(0.8, 0.2))
loaded
```


## Tossing a coin

Once you have defined a `"coin"` object, you can use the function `toss()` to
_toss_ a coin a given number of `times`. The output will be an object of class 
`"toss"`, which will contain the vector of `tosses`:

```{r}
toss500 <- toss(mycoin, times = 100)
toss500
```

The `summary()` function gives you a nice table of frequencies:
```{r}
summary(toss500)
```


## Plot of relative frequencies

You can use the function `plot()` to create a graohic with the relative 
frequencies of a series of coin tosses:

```{r, fig.show='hold'}
plot(toss500)
```
